---
title: Bioinformatic Pipeline - "Hiding in Plain Sight - Genome-wide recombination
  and horizontal gene transfer drive clonal diversity in Fusarium oxysporum fsp ciceris"
author: "Fayyaz A, Robinson G, Chang PL, Bekele D, Yimer SM, Carrasquilla-Garcia N, Negash K, Anand K. Surendrarao, Eric J.B. von Wettberg, Robbertse B, Seid Ahmedi, Tesfaye K, Fikre Ab, Farmer AD and Cook DR"
#date: '2023-02-07'
output:
  #rmdformats::readthedown:
  #  toc_depth: 5
  #  self_contained: true
  #  thumbnails: false
  #  lightbox: true
  #  gallery: true
  #  highlight: tango
  html_document: default
  #pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(reticulate)
```

---

## Contents

### 1. [Overview](#Overview)
### 2. [Genome Assembly](#assembly)
### 3. [Gene Annotation and Analyses](#gene_annotation)
### 4. [Identification of SNP and Genetic Groups](#read_mapping)
### 5. [Population Analyses](#pop_gen)
### 6. [References](#refs)

---

## Help with Rmarkdown
### https://holtzy.github.io/Pimp-my-rmd/
### https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

## <a id="Overview"></a>1. Overview

Plot overview pipeline

## <a id="assembly"></a>2. Genome Assembly

```{r, eval=F}
#####################################################################################
######################### Need input from Amna/Peter ################################
#####################################################################################
```

A. Quality check with FastQC
B. Low quality reads and adapters trimmed using Trimmomatic v36
C. Error correction using ALLPATHS-LG
D. Assembly using SPAdes
E. Genome completeness assessed using BUSCO v3 (Ascomycota odb9 dataset)

## <a id="gene_annotation"></a>3. Gene Annotation and Analyses

```{r, eval=F}
#####################################################################################
######################### Need input from Amna/Peter ################################
#####################################################################################
```

A. Average nucleotide identity calculated with Pyani
```{bash, eval=F}
/home/localhost/adf/sw/bin/singularity run /erdos/adf/sw/singularityImages/average_nucleotide_identity-v0.2.9.simg -i all-file-pyani/ -g -m ANIb --workers 20 -o pyani-result-552-genes
```
B. Annotation
  - RepeatModeler
  - MAKER
  - Interpro, GeneOntology, AHRD
  - Gene Enrichment with BLAST2GO
  - SIX genes homologs BLASTp
  - Orthologs identified with Orthofinder (default excpt with -msa)
C. Phylogenetic Analyses
  - BEAST
  - iTOL
  - Robinson-Foults with ETE3 toolkit
D. Pangenome
  - PanOCT

## <a id="read_mapping"></a>4. Identification of SNP and genetic groups

### A. Read mapping

### i. Illumina read mapping was conducted with BWA MEM 0.7.9a-r786

Example -- https://www.illumina.com/products/by-type/informatics-products/basespace-sequence-hub/apps/bwa-aligner.html

Tools required:

- BWA MEM v0.7.9a-r786

- SAMtools v1.3.1

- Picard 

```{bash, eval=FALSE}
#create index for F. oxysporum reference genome (Fol4287 - GCF_000149955.1)
bwa index [-a bwtsw|is] reference.fasta index_prefix

#running BWA (looped through all Illumina paired-end read data)
#loops through Illumina read prefixes 
for f in $(ls SRR* | cut -d"_" -f1); \
do bwa mem -M -t 32 reference.fasta -P ${f}_1.fastq ${f}_2.fastq > ${f}.sam; \
done

#converts SAM file to BAM file
for f in $(ls *.sam | cut -d"." -f1); do samtools fixmate -O bam ${f}.sam ${f}.bam
```

### ii. Variant calling with GATK v4.1 against Fol4287 reference

Tools required:

- GATK v4.1

```{bash, eval=F}
#####################################################################################
######################### Need input from Amna/Peter ################################
#####################################################################################
```


### B. Filtering VCF file 

Tools required:

bcftools/1.16

vcftools/b240116

mummer/3.23

For most population analyses, we exclusively used high quality genomes (based on BUSCO completeness > 97%, N50, number of scaffolds and genome size) - see "Genome Assembly". This provided a total list of 120 genomes, 99 of which were taxonomically within the FOSC (Fig. 2). This is reduced from the earlier VCF file, which contained 320 isolates (Supplemental Fig. 1)

Reducing the VCF file from 320 to 120, was done using bcftools (custom bash script).

### i. We split the VCF into separate single isolate VCF files

```{bash, eval=F}
######## ./splitting_vcfs_retaining.sh ###########

#!/usr/bin/env bash
for file in *.vcf.gz; do
  for sample in `bcftools query -l $file`; do
    bcftools view -Oz -s $sample -o ${file/.vcf*/.$sample.vcf.gz} $file
  done
done
```

### ii. We indexed all the single isolate VCF files

```{bash, eval=F}
for f in *vcf.gz; do tabix -f -p vcf $f; done
```

### iii. We merged isolates from list of high quality genomes (information obtained from assembly data)

```{bash, eval=F}
bcftools merge --file-list isolate_list > ethiopia_fo_structure.vcf
```

### iv. We removed heterozygous/ambiguous SNPs using sed

```{bash, eval=F}
sed -i 's#0/1|0/2|0/3|0/4|1/2|1/3|1/4|2/3|2/4|3/4#[.]/[.]#g' ethiopia_fo_structure.vcf
```

### v. We removed SNPs with high missingness and evaluated MAF filtering

To ensure that the subsequent VCF was high quality, especially for linkage disequilibrium, we adjusted filters to learn how different filters affect the subsequent linkage disequilibrium analyses. 

The main two parameters were missingness and minor allele frequency (MAF). MAF was of particular importance to us, as often researchers adjust MAF cutoff as >0.05 without regard for the effects on the population. 

```{bash, eval=F}
#VCF file for STRUCTURE groupings
#Note that we kept the MAF filter for the STRUCTURE groupings only because it is generally recommended.
vcftools --vcf ethiopia_fo_structure.vcf --maf 0.05 --max-missing 0.9 --recode --stdout | bgzip -c > all_ethiopia_isolates.0_05.vcf.gz 

############## Add that we pruned the data using PLINK too ###############

#total number of SNPs = 74668

#VCF file for population genetic analyses
vcftools --vcf ethiopia_fo_structure.vcf --max-missing 0.9 --recode --stdout | bgzip -c > all_ethiopia_isolates.vcf.gz #total number of SNPs = 136772

#note that this was also conducted but there were only minor differences in the outcomes. The overall clonal groupings remained the same. The LD decay curve showed minor differences

#include plot /home/guy/Documents/Research/Cook/fusarium_pnas_final/old/Linkage/ld_total_maf.png
```

### vi. To ensure SNPs were core components of the genome, mapping was informed by additional whole genome alignments (MUMmer)

This was done on per chromosome basis (Fol4287 1-15)

```{bash, eval=F}
#Here is the example of chromosome 7 (Suppl. Fig. 2)
for f in Fo-Et-0*; do nucmer -maxmatch -c 100 -p ${f}_chr7 chr7.fna ${f}; done #outputs delta file

# Whole genome alignments were visualized and filtered in R
Rscript ./mummer_coverage.R

# This led to the removal of any SNPs mapped to chromosomes 3, 6, 14 and 15 as well as supercontig 27 (end of chromosome 1)
# Any reference contigs that mapped poorly to the Illumina genomes were removed running awk itieratively
# e.g. awk '($0 !~ "Supercontig_2.27"){print}' all_ethiopia_isolates.vcf.gz
```


### C. Identification of clonal groups using STRUCTURE

Tools required:

structure-console/2.3.4

structure harvester (https://taylor0.biology.ucla.edu/structureHarvester/)

CLUMPP/1.1.2

Subdividing clonal lineages in fungal populations remains challenging. Traditionally, reasearchers have determined clonality based on identical MLGs. This approach fails on a whole genome scale due to the accumulation of erroneous mutations. We employed two different methods to subdivide the lineages: phylogenetics (see Gene Annotation and Analyses - BEAST) and K clustering (via STRUCTURE). 

### i. Random subsampling SNPs

To avoid SNP selection bias, the VCF file was randomly subsampled at a rate of ~5%. 

```{bash, eval=F}
# Range(number of SNPs):
# Subsample 1: 3756
# Subsample 2: 3665
# Subsample 3: 3826
# Subsample 4: 3653
# Subsample 5: 3764
# Subsample 6: 3777
# Subsample 7: 3682
# Subsample 8: 3744
# Subsample 9: 3784
# Subsample 10: 3747


bgzip -dc all_ethiopia_isolates.vcf.gz | awk 'BEGIN {srand()} !/^$/ { if (rand() <= .1) print $0}' | bgzip -c > ethiopia_fo_structure.0_05.vcf.gz_sub1.vcf.gz

bgzip -dc all_ethiopia_isolates.vcf.gz | awk 'BEGIN {srand()} !/^$/ { if (rand() <= .1) print $0}' | bgzip -c > ethiopia_fo_structure.0_05.vcf.gz_sub2.vcf.gz

bgzip -dc all_ethiopia_isolates.vcf.gz | awk 'BEGIN {srand()} !/^$/ { if (rand() <= .1) print $0}' | bgzip -c > ethiopia_fo_structure.0_05.vcf.gz_sub3.vcf.gz

bgzip -dc all_ethiopia_isolates.vcf.gz | awk 'BEGIN {srand()} !/^$/ { if (rand() <= .1) print $0}' | bgzip -c > ethiopia_fo_structure.0_05.vcf.gz_sub4.vcf.gz

#etc.
```

VCF subsample files were converted into genotype files using a manual bash script

```{bash, eval=F}
############ ./convertvcftogeno.sh #########
#!/usr/bin/env bash
file=$1
out=${file}.out
bgzip -dc ${file} | awk '$1 ~ "Super" || $1 ~ "#CHROM" {print}' | awk '$1 !~ "##" {print}' | sed 's#0/0#0#g;s#1/1#1#g;s#2/2#2#g;s#3/3#3#g;s#4/4#4#g;s#0/1#-9#g;s#0/2#-9#g;s#0/3#-9#g;s#0/4#-9#g;s#1/2#-9#g;s#1/3#-9#g;s#1/4#-9#g;s#2/3#-9#g;s#2/4#-9#g;s#3/4#-9#g;s#./.#-9#g' | cut -f 2,10- | awk '
{ 
    for (i=1; i<=NF; i++)  {
        a[NR,i] = $i
    }
}
NF>p { p = NF }
END {    
    for(j=1; j<=p; j++) {
        str=a[1,j]
        for(i=2; i<=NR; i++){
            str=str" "a[i,j];
        }
        print str
    }
}' > $out
```

### ii. STRUCTURE
  
Ran STRUCTURE on each subsampling 10 times: burnin = 10,000 and MCMC reps = 50,000. K vals = 2-20. Check out Mainparams and Extraparams for additional technical information.
  
```{bash, eval=F}
# e.g.
#!/bin/bash
module load structure-console/2.3.4
structure -K 10 -o structure_large10-1-2
```

### iii. STRUCTUREharvester

Using outputs of STRUCTURE, combine them using STRUCTURE_HARVESTER (online web server: http://taylor0.biology.ucla.edu/structureHarvester/) to produce .indfiles for CLUMPP.

### iv. CLUMPP

Ran CLUMPP on HPC cluster to collate 10 permutations of each subsample. Used LargeKGreedy algorithm.

```{bash, eval=F}
#e.g.
#!/bin/bash
module load CLUMPP/1.1.2
CLUMPP K2.paramfile -i K12.indfile -k 12 -c 133 -r 10 -m 3 -w 0 -s 2 -o K12.outfile -j K12.miscfile
```

### v. STRUCTURE plotted with R script

```{bash, eval=F}
######### ./structure_plotting_larger_collection #############
```

### Clonal groups were assigned based on Euclidean distances

```{bash, eval=F}
######### ./plotting_structure_network.R ##########
./plotting_structure_network.R --list "list of all .outfile paths" --label "list of isolate names (in order)" --title "title of graph (usually K val included)"
```



## <a id="pop_gen"></a>5. Population Analyses

A. Geographic Analyses
  - Plotting onto maps in R
  - Mantel tests
B. Linkage disequilibrium
  - PLINK (note multiallelic SNPs automatically filtered out)
  - Additional filtering for derived SNPs and ancestral SNPs
  /media/guy/Guy HardDrive/vcf_individuals_largerCollection/final_ld/clades
    - BASH scripts
  - LD with r2 (PLINK)
    - Inter and intra-chromosomal LD
    - Genome-wide LD across genome
  - LD with rbarD (Poppr)
C. Nucleotide Diversity (pi)
D. Tajima's D
E. Fst (Weighted)
  - Core SNP Fst
  - Pangenome Fst
  - BUSCO Fst

## <a id="refs"></a>6. References